---
title: "DSBA Group Project"
author: "Ethan Pinto"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The data set faces.RData (courtesy of Havard Rue, NTNU) contains portraits or grey
level images of n = 200 medical students at Stanford University, where i = 1,…,100 are
male portraits and i = 101,…, 200 are female portraits. Each portrait can be thought of
as a matrix of 100 by 100 pixels and is recorded as a vector of length 𝑝 = 100
2 = 10000,
treating each pixel as a variable. The function image() on R scales and maps each pixel to a grey
value (black being '0' and white being '1') and can be used to display a 10000 × 1 vector as an
image.

Project setup:
```{r}
load("faces.rdata")

# install.packages("pls")
# install.packages("rda")
# install.packages("klaR")

library(pls)
library(MASS)
library(sda)
library(rda)
library(klaR)
```


a) Estimate the mean of the portraits for each gende and display them as 100 × 100
images. Does the mean vary across the image? Which features are represented in the
estimated mean portraits? The portraits are a mix of images taken at the shoulders and
at the chin, such that the relative size of the face varies significantly. The variable
shoulder indicates whether a portrait includes the shoulders or not.

```{r}
reshape_portrait <- function(vector) {
  matrix(vector, nrow = 100, ncol = 100)
}

mean_male_portrait <- apply(faces[, 1:100], 1, mean)
mean_female_portrait <- apply(faces[, 101:200], 1, mean)

mean_male_image <- reshape_portrait(mean_male_portrait)
mean_female_image <- reshape_portrait(mean_female_portrait)

image(1:100, 1:100, mean_male_image[, 100:1], col = grey(0:255/255), main = "Mean Male Portrait")
image(1:100, 1:100, mean_female_image[, 100:1], col = grey(0:255/255), main = "Mean Female Portrait")


```

b) Find the principal components of the whole data set and display the first three eigenvectors (the
principal component directions) as 100 × 100 images. The images of the eigenvector are
usually referred to as eigenfaces in context of face recognition. What do the first three
components represent in terms of facial or image features? How many principal components are
needed to express 80% of the variation in the data set?

```{r}
pca <- prcomp(faces, scale. = FALSE)
eigenvectors <- pca$x[, 1:3]


eigenmatrix1 <- matrix(eigenvectors[,1],nrow=100, ncol=100)

eigenmatrix2 <- matrix(eigenvectors[,2],nrow=100, ncol=100)

eigenmatrix3 <- matrix(eigenvectors[,3],nrow=100, ncol=100)

image(eigenmatrix1, col = gray.colors(256))
image(eigenmatrix2, col = gray.colors(256))
image(eigenmatrix3, col = gray.colors(256))

summary(pca)
```

c) Find the first three principal components (𝑛 × 1 vectors) of the whole data set. Plot the first
and second principal components and the first and third principal components against each
other and color the observations according to gender. Then color the plots according to the
shoulders being present or not. What do you conclude? Display all four figures.

```{r}
pc1 <- pca$x[, 1]
pc2 <- pca$x[, 2]
pc3 <- pca$x[, 3]

gender <- c(rep("Male", 100), rep("Female", 100))


plot(pc1, pc2, col = ifelse(gender == "Male", "blue", "red"), xlab = "PC1", ylab = "PC2", main = "PC1 vs PC2 by Gender")
plot(pc1, pc3, col = ifelse(gender == "Male", "blue", "red"), xlab = "PC1", ylab = "PC3", main = "PC1 vs PC3 by Gender")

plot(pc1, pc2, col = ifelse(shoulder, "green", "purple"), xlab = "PC1", ylab = "PC2", main = "PC1 vs PC2 by Shoulders")
plot(pc1, pc3, col = ifelse(shoulder, "green", "purple"), xlab = "PC1", ylab = "PC3", main = "PC1 vs PC3 by Shoulders")

```

d) Recode gender as an indicator variable
        −1, if Gi = male
yi = {
        1, if Gi = female

Construct a classifier for gender using principal component regression (PCR) with 𝑦𝑖 as
the response, and classify an observation as female if f(xi) > 0. Use leave-one-out
cross-validation to select the number of components, 𝑚 < 200, and select the smallest
m if the cross-validation error has several minima. What is the misclassification rate (the
error with a 1-0 loss function on the training data)?

```{r}
y <- ifelse(gender == "Male", -1, 1)

faces_df <- data.frame(faces)
faces_df <- cbind(y = y, faces_df)

set.seed(6115) 
pcr_model <- pcr(y ~ ., data = faces_df, scale = TRUE, validation = "LOO")

validation_stats <- summary(pcr_model)
opt_m <- which.min(validation_stats$val$MSEP)

predictions <- predict(pcr_model, ncomp = opt_m)
predicted_class <- ifelse(predictions > 0, 1, -1)
misclassification_rate <- mean(predicted_class != y)

print(misclassification_rate)

```

e) Construct a classifier for gender using partial least square (PLS) with 𝑦𝑖 as the response,
and classify an observation as female if 𝑓̂(𝑥𝑖) > 0. Use leave-one-out cross-validation
to select the number of PLS directions, m < 200, and select the smallest m if the crossvalidation error has several minima. What is the misclassification rate (the error with a
1-0 loss function on the training data) now? Why does the optimal m for PCR and PLS
differ

```{r}
set.seed(6115)
pls_model <- plsr(y ~ ., data = faces_df, scale = TRUE, validation = "LOO")

validation_stats_pls <- summary(pls_model)
opt_m_pls <- which.min(validation_stats_pls$val$MSEP)

predictions_pls <- predict(pls_model, ncomp = opt_m_pls)
predicted_class_pls <- ifelse(predictions_pls > 0, 1, -1)
misclassification_rate_pls <- mean(predicted_class_pls != y)

print(misclassification_rate_pls)

```

Ordinary linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) cannot be
used in a high-dimensional setting with p > n, as the sample covariance matrix will be singular.
You will now explore two ways to avoid this issue; reduce the dimension and regularize the
covariance matrix.

f) Let the whole data set (10000 × 200) to be represented by the first five principal
components (5 × 200). Use the reduced data as input for QDA and construct a classifier for
gender. What is misclassification rate on the training data? Plot the quadratic decision
boundaries in the figures from 3(c).

```{r}
faces_reduced <- as.data.frame(pca$rotation[, 1:5])

y_factor <- as.factor(ifelse(1:200 <= 100, "Male", "Female"))
qda_model <- qda(x = faces_reduced, grouping = y_factor)

qda_predictions <- predict(qda_model, faces_reduced)
misclassification_rate_qda <- mean(qda_predictions$class != y_factor)

print(misclassification_rate_qda)

mean_pc3 <- mean(faces_reduced[, 3])
mean_pc4 <- mean(faces_reduced[, 4])
mean_pc5 <- mean(faces_reduced[, 5])

xrange <- range(faces_reduced[, 1])
yrange <- range(faces_reduced[, 2])

grid <- expand.grid(PC1 = seq(from = xrange[1], to = xrange[2], length = 100),
                    PC2 = seq(from = yrange[1], to = yrange[2], length = 100),
                    PC3 = mean_pc3,
                    PC4 = mean_pc4,
                    PC5 = mean_pc5)

grid$predicted_class <- predict(qda_model, newdata = grid)$class

plot(faces_reduced[, 1], faces_reduced[, 2], col = as.numeric(y_factor), xlab = "PC1", ylab = "PC2", pch = 19)
plot(faces_reduced[, 1], faces_reduced[, 3], col = as.numeric(y_factor), xlab = "PC1", ylab = "PC3", pch = 19)

contour(xrange, yrange, matrix(as.numeric(grid$predicted_class), length(xrange), length(yrange)), add = TRUE)

```

g) Now consider the gender and the shoulder categories together as four different categories
(maleShoulder, maleNoShoulder, femaleShoulder and femaleNoShoulder). Use the first five
principal components (5_200) to construct a QDA classifier for the four gender-shoulder
categories and classify the training data. Merge the classifications for the groups with and
without shoulders within each gender, such that you end up with a classification for gender.
Compare the result to the classification in f). Does the misclassification rate decrease when
taking into account the shoulders?

```{r}
gender_shoulder <- ifelse(y == -1, ifelse(shoulder, "maleShoulder", "maleNoShoulder"),
                          ifelse(shoulder, "femaleShoulder", "femaleNoShoulder"))

qda_model_gs <- qda(x = faces_reduced, grouping = as.factor(gender_shoulder))
qda_predictions_gs <- predict(qda_model_gs, faces_reduced)
misclassification_rate_gs <- mean(qda_predictions_gs$class != gender_shoulder)
print(misclassification_rate_gs)

```

h) Finally, return to the whole data set and use a version of regularized LDA to classify only
gender. Motivate how you regularize the sample covariance matrix and argue for your
choice of tuning (penalty) parameter. What is the misclassification rate?

(This method takes a very long time)

```{r}
lda_data <- as.data.frame(faces_pca)
lda_data$y_factor <- y_factor

lda_model <- lda(y_factor ~ ., data = lda_data)

lda_predictions <- predict(lda_model, newdata = lda_data)
predicted_gender <- as.numeric(lda_predictions$class)
misclassification_rate_lda <- mean(predicted_gender != as.numeric(lda_data$y_factor))

print(misclassification_rate_lda)

```

Faster method:

```{r}
rda_model <- rda(x=data.matrix(faces),y=y)
rda_predictions <- predict(rda_model, x=data.matrix(faces),y=y,xnew=data.matrix(faces),alpha=0, delta=0.667)

misclassification_rate_rda <- mean(rda_predictions != y)

print(misclassification_rate_rda)
```

